On joint estimation of ancestral states and phylogenetic model parameters

### 2017-10-16 joint_inf.tex
Cleaning out addressed notes.

Now substitute in values for $p_{\emptyset}$ and $p_{13}$ from Table~\ref{tab:sitepatprob}.
%EM Not sure what's being suppressed here. Sounds naughty.
%das: I meant suppressing terms not involving $w'$, but I guess I already did that...
Simplifying, the likelihood can be written
$$
\ell(w') = \alpha_1\log(\beta+\gamma w') + \alpha_2\log(\beta-\gamma w') + \log(1+w')
$$
where
$$
\alpha_1 := \alpha_1(x, y) = 1+x^2+y^2+4xy^2+x^2y^2,
$$
$$
\alpha_2 := \alpha_2(x, y) = 1+x^2+y^2-4xy^2+x^2y^2,
$$
$$
\beta := \beta(x, y) = 1+x^2+y^2+x^2y^2,
$$
$$
\gamma := \gamma(x, y) = 4xy.
$$
Since $x$ and $y$ fall between zero and one, we make use of the inequalities (excepting the cases of $x=y=0$ and $x=y=1$)
$$
\alpha_1 > \alpha_2
$$
and
$$
\beta > \gamma.
$$
% Not sure if I need to show this, but if x and y aren't 1 it's 1+x^2+y^2+x^2y^2 > 4xy => (1+x^2)(1+y^2) > 4xy and 1+x^2 > 2x, 1+y^2 > 2y qed
%EM Rather confusing to be using ' for two things here. What about using a dot over the l? In retrospect something other than ' would have been good because of the proliferation of w^{'2} etc. Prob not worth changing now though.
The derivative of the log likelihood is
$$
\ell'(w') := \frac{d}{dw'} \ell(w') = \frac{\alpha_1 \gamma}{\beta+\gamma w'} - \frac{\alpha_2 \gamma}{\beta-\gamma w'} + \frac{1}{1+w'}.
$$
%EM I don't think you mean "any inequality"... what about sin(ell') > 0?
The inequality $\beta > \gamma$ implies that this function stays finite and that, when considering $\ell'(w') \lessgtr 0$, we equivalently consider $f(w') \lessgtr 0$ where $f$ is the quadratic function
\begin{align*}
f(w') &= (w')^2\cdot(-\gamma^2\alpha_1-\gamma^2\alpha_2-\gamma^2) \\
      &\qquad + w'\cdot(\gamma\alpha_1\beta-\gamma^2\alpha_1-\gamma\alpha_2\beta-\gamma^2\alpha_2) \\
      &\qquad + (\gamma\alpha_1\beta-\gamma\alpha_2\beta+\beta^2).
\end{align*}
%EM Is this obvious at this point? Couldn't b^2 - 4ac be zero in principle?
%das: only if x or y is zero (though one could be nonzero).
%b^2 - 4ac is always \ge 0 by \alpha_1 > \alpha_2, so b^2 - 4ac can only be zero if b is zero and a or c is zero.
%c is never zero, so this means a must be zero for this to happen, which only occurs if \gamma = 0, i.e., if x or y is zero.
%Added clarification.
Excepting all the cases where $x=0$ or where $y=0$, this implies $\ell'$ has two zeros according to the quadratic formula.
Because $\alpha_1 > \alpha_2$ we have $\ell'(0) > 0$, and thus $\ell$ is increasing at $w'=0$.
This implies that if the smaller of the zeros of $f(w')$ is greater than one, then $\hat{w}' \equiv 1$.
Using the quadratic formula with
$$
a(x, y) = -\gamma^2 \alpha_1 - \gamma^2 \alpha_2 - \gamma^2,
$$
$$
b(x, y) = \gamma \alpha_1 \beta - \gamma^2\alpha_1 - \gamma \alpha_2 \beta - \gamma^2 \alpha_2,
$$
$$
c(x, y) = \gamma \alpha_1 \beta - \gamma \alpha_2 \beta + \beta^2,
$$
the smaller zero is
$$
\hat{w}'(x, y) = \frac{-b(x, y) - \sqrt{b(x, y)^2 - 4a(x, y)c(x, y)}}{2a(x, y)},
$$
which is a function of the generating parameters $x$ and $y$.
We plot the region where $\hat{w}'(x, y) \ge 1$ in Figure~\ref{fig:bl-inconsistency}.

@@

Our plan of attack is to determine inequalities at which a particular form of likelihood is necessary and plot the regions of $x$ and $y$ where these occur in addition to the region where $\hat{w}'$ is estimated to be one.
Since we are interested in the region where $\hat{w}'=1$, we choose either the term $(1-x')^2(1+w')(1+y')^2$ or $(1+x')^2(1+w')(1-y')^2$ in Table~\ref{tab:likelihoods} for $\siteSplit_j=\{1,3\}$.
%EM I don't think I follow the thread of argument here-- what's being assumed when? I think you are assuming $\hat{y}' > \hat{x}'$ ?
Now, assume that, having estimated $\hat{x}'$ and $\hat{y}'$, $\hat{y}' > \hat{x}'$.
In this case,
$$
(1-\hat{x}'^2)(1+\hat{y}'^2) > (1+\hat{x}'^2)(1-\hat{y}'^2),
$$
so we use the likelihood
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
    &=        p_{\emptyset}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
    &\qquad + p_{1}          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \\
    &\qquad + p_{2}          \cdot\log(1+x'^2-y'^2-x'^2y'^2) \\
    &\qquad + p_{3}          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \\
    &\qquad + p_{12}         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \\
    &\qquad + p_{13}         \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
    &\qquad + p_{23}         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \\
    &\qquad + p_{123}        \cdot\log(1+x'^2-y'^2-x'^2y'^2) \\
    &\qquad - \log(8) \\
    &\qquad + p_{\emptyset}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \\
    &\qquad + p_{1}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
    &\qquad + p_{2}          \cdot\log((1+x')^2   (1+w')(1+y')(1-y')) \\
    &\qquad + p_{3}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
    &\qquad + p_{12}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \\
    &\qquad + p_{13}         \cdot\log((1-x')^2   (1+w')(1+y')^2) \\
    &\qquad + p_{23}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \\
    &\qquad + p_{123}        \cdot\log((1+x')^2   (1+w')(1+y')(1-y')).
\end{align*}
This simplifies to
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
    &=        p_{\emptyset}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
    &\qquad + p_{13}         \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
    &\qquad + (p_{1} + p_{3})          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \\
    &\qquad + (p_{2} + p_{123})          \cdot\log(1+x'^2-y'^2-x'^2y'^2) \\
    &\qquad + (p_{12} + p_{23})         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \\
    &\qquad + (2-p_{1}-p_{3}-p_{12}-p_{23}-2p_{13})\log(1+x') \\
    &\qquad + (p_{1}+p_{3}+p_{12}+p_{23}+2p_{13})\log(1-x') \\
    &\qquad + (2-p_{2}-p_{12}-p_{23}-p_{123})\log(1+y') \\
    &\qquad + (p_{2}+p_{12}+p_{23}+p_{123})\log(1-y') \\
    &\qquad + \log(1+w').
\end{align*}
We need to obtain
$$
(\hat{x}', \hat{y}', \hat{w}') = \arg\max_{x, y, w} \ \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
$$
as functions of $x$ and $y$ and verify the region where $\hat{w}'(x, y) = 1$ and $\hat{y}'(x, y) > \hat{x}'(x, y)$.
This region will be consistent with one always estimating $\hat{w}'$ to be one.
This can similarly be done for the other likelihood term with $\hat{x}'(x, y) > \hat{y}'(x, y)$.
To do is to find these functions---or subsets of their varieties---in as closed a form as possible.

>>>>

Now substitute in values for $p_{\emptyset}$ and $p_{13}$ from Table~\ref{tab:sitepatprob}.
Simplifying, the likelihood can be written
$$
\ell(w') = \alpha_1\log(\beta+\gamma w') + \alpha_2\log(\beta-\gamma w') + \log(1+w')
$$
where
$$
\alpha_1 := \alpha_1(x, y) = 1+x^2+y^2+4xy^2+x^2y^2,
$$
$$
\alpha_2 := \alpha_2(x, y) = 1+x^2+y^2-4xy^2+x^2y^2,
$$
$$
\beta := \beta(x, y) = 1+x^2+y^2+x^2y^2,
$$
$$
\gamma := \gamma(x, y) = 4xy.
$$
Since $x$ and $y$ fall between zero and one, we make use of the inequalities (excepting the cases of $x=y=0$ and $x=y=1$)
$$
\alpha_1 > \alpha_2
$$
and
$$
\beta > \gamma.
$$
% Not sure if I need to show this, but if x and y aren't 1 it's 1+x^2+y^2+x^2y^2 > 4xy => (1+x^2)(1+y^2) > 4xy and 1+x^2 > 2x, 1+y^2 > 2y qed
%EM Rather confusing to be using ' for two things here. What about using a dot over the l? In retrospect something other than ' would have been good because of the proliferation of w^{'2} etc. Prob not worth changing now though.
The derivative of the log likelihood is
$$
\ell'(w') := \frac{d}{dw'} \ell(w') = \frac{\alpha_1 \gamma}{\beta+\gamma w'} - \frac{\alpha_2 \gamma}{\beta-\gamma w'} + \frac{1}{1+w'}.
$$
The inequality $\beta > \gamma$ implies that this function stays finite and that, when considering $\ell'(w') \lessgtr 0$, we equivalently consider $f(w') \lessgtr 0$ where $f$ is the quadratic function
\begin{align*}
f(w') &= (w')^2\cdot(-\gamma^2\alpha_1-\gamma^2\alpha_2-\gamma^2) \\
      &\qquad + w'\cdot(\gamma\alpha_1\beta-\gamma^2\alpha_1-\gamma\alpha_2\beta-\gamma^2\alpha_2) \\
      &\qquad + (\gamma\alpha_1\beta-\gamma\alpha_2\beta+\beta^2).
\end{align*}
% Note from das: b^2 - 4ac is always \ge 0 by \alpha_1 > \alpha_2, so b^2 - 4ac can only be zero if b is zero and a or c is zero. c is never zero, so this means a must be zero for this to happen, which only occurs if \gamma = 0, i.e., if x or y is zero.
Excepting all the cases where $x=0$ or where $y=0$, this implies $\ell'$ has two zeros according to the quadratic formula.
Because $\alpha_1 > \alpha_2$ we have $\ell'(0) > 0$, and thus $\ell$ is increasing at $w'=0$.
This implies that if the smaller of the zeros of $f(w')$ is greater than one, then $\hat{w}' \equiv 1$.
Using the quadratic formula with
$$
a(x, y) = -\gamma^2 \alpha_1 - \gamma^2 \alpha_2 - \gamma^2,
$$
$$
b(x, y) = \gamma \alpha_1 \beta - \gamma^2\alpha_1 - \gamma \alpha_2 \beta - \gamma^2 \alpha_2,
$$
$$
c(x, y) = \gamma \alpha_1 \beta - \gamma \alpha_2 \beta + \beta^2,
$$
the smaller zero is
$$
\hat{w}'(x, y) = \frac{-b(x, y) - \sqrt{b(x, y)^2 - 4a(x, y)c(x, y)}}{2a(x, y)},
$$
which is a function of the generating parameters $x$ and $y$.
We plot the region where $\hat{w}'(x, y) \ge 1$ in Figure~\ref{fig:bl-inconsistency}.

@@

Our plan of attack is to determine inequalities at which a particular form of likelihood is necessary and plot the regions of $x$ and $y$ where these occur in addition to the region where $\hat{w}'$ is estimated to be one.
Since we are interested in the region where $\hat{w}'=1$, we choose either the term $(1-x')^2(1+w')(1+y')^2$ or $(1+x')^2(1+w')(1-y')^2$ in Table~\ref{tab:likelihoods} for $\siteSplit_j=\{1,3\}$.
Now, assume that, having estimated $\hat{x}'$ and $\hat{y}'$, $\hat{y}' > \hat{x}'$.
In this case,
$$
(1-\hat{x}'^2)(1+\hat{y}'^2) > (1+\hat{x}'^2)(1-\hat{y}'^2),
$$
so we use the likelihood
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
    &=        p_{\emptyset}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
    &\qquad + p_{1}          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \\
    &\qquad + p_{2}          \cdot\log(1+x'^2-y'^2-x'^2y'^2) \\
    &\qquad + p_{3}          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \\
    &\qquad + p_{12}         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \\
    &\qquad + p_{13}         \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
    &\qquad + p_{23}         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \\
    &\qquad + p_{123}        \cdot\log(1+x'^2-y'^2-x'^2y'^2) \\
    &\qquad - \log(8) \\
    &\qquad + p_{\emptyset}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \\
    &\qquad + p_{1}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
    &\qquad + p_{2}          \cdot\log((1+x')^2   (1+w')(1+y')(1-y')) \\
    &\qquad + p_{3}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
    &\qquad + p_{12}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \\
    &\qquad + p_{13}         \cdot\log((1-x')^2   (1+w')(1+y')^2) \\
    &\qquad + p_{23}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \\
    &\qquad + p_{123}        \cdot\log((1+x')^2   (1+w')(1+y')(1-y')).
\end{align*}
This simplifies to
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
    &=        p_{\emptyset}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
    &\qquad + p_{13}         \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
    &\qquad + (p_{1} + p_{3})          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \\
    &\qquad + (p_{2} + p_{123})          \cdot\log(1+x'^2-y'^2-x'^2y'^2) \\
    &\qquad + (p_{12} + p_{23})         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \\
    &\qquad + (2-p_{1}-p_{3}-p_{12}-p_{23}-2p_{13})\log(1+x') \\
    &\qquad + (p_{1}+p_{3}+p_{12}+p_{23}+2p_{13})\log(1-x') \\
    &\qquad + (2-p_{2}-p_{12}-p_{23}-p_{123})\log(1+y') \\
    &\qquad + (p_{2}+p_{12}+p_{23}+p_{123})\log(1-y') \\
    &\qquad + \log(1+w').
\end{align*}
We need to obtain
$$
(\hat{x}', \hat{y}', \hat{w}') = \arg\max_{x, y, w} \ \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
$$
as functions of $x$ and $y$ and verify the region where $\hat{w}'(x, y) = 1$ and $\hat{y}'(x, y) > \hat{x}'(x, y)$.
This region will be consistent with one always estimating $\hat{w}'$ to be one.
This can similarly be done for the other likelihood term with $\hat{x}'(x, y) > \hat{y}'(x, y)$.
To do is to find these functions---or subsets of their varieties---in as closed a form as possible.
